Data Specialization for a Staged Lambda Calculus

### 2014-11-05 paper/semantics.tex
few additions to splitting
lots of easy changes in 3 and 4

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
2`prev{`
  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
}`
\end{lstlisting}
The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
This search process is called {\em speculation}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
2`prev{`1`
  next{`2` 
    prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
  `1`}`2`
}`
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
2`(f 20) * (f 20) * 7`
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

@@

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`let x = (next{`2`yhat`1`}, 7) in
  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
}`
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`next{`2`
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
  `1`}`2`
}`
\end{lstlisting}
We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
    `1`next{`2`yhat * yhat * 7`1`}
`2`}`
\end{lstlisting}
Once again, we lift the results of speculation into a substitution:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
[zhat|->yhat * yhat * 7]
    `1`next{`2`zhat`1`}`2`
}`
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

@@

\subsubsection{Evaluation and Speculation}
\label{ssec:dynamics}

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (future context).
These judgments are related as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.
In order to avoid duplicating subterms, evaluation collects explicit
substitutions into a single {\em future context} that is also the output of
evaluation.
\item Speculation sends a stage~\bbtwo\ term to a residual.
\item Reification sends a residual and a future context to another residual.  
\item Evaluation depends on speculation for $\next$ terms, which have
stage~\bbtwo\ subterms.
\item Speculation depends on evaluation for $\prev$ and $\pause$, which have
stage~\bbone\ subterms. The future context produced by evaluation is also
reified here.
\end{itemize}

@@

\ur{I would recommend using ``1-value'' and ``2-value'' instead of
  partial values and residuals.  The former are more accurate, more
  direct, more succinct, and does not lead to confusions with other
  residuals.}

\ur{Many incorrectly hyphenated ``partial values'' in this text.}

The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
\bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a {\em future environment} $\xi$ and a partial value $v$. The future
environment implements the explicit substitutions in our example---it maps fresh
stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
residuals. For non-staging features of \lang, $\redonesym$ is essentially
standard call-by-value evaluation, and gathers the subterms' future environments
into a single one.

@@

At $\next~e$, in contrast, $\redonesym$ speculates into $e$ for stage~\bbone\ 
subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.
Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
environment which maps that variable to the residual ($\hat y \mapsto q$).

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
1`next{
  `2`if true 
  then hold{`1`3+4`2`} 
  else prev{`1`spin() (* loops forever *)`2`}`1`
}`
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

>>>>

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
2`prev{`
  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
}`
\end{lstlisting}
The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
This search process is called {\em specialization}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
2`prev{`1`
  next{`2` 
    prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
  `1`}`2`
}`
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
2`(f 20) * (f 20) * 7`
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

@@

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`let x = (next{`2`yhat`1`}, 7) in
  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
}`
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`next{`2`
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
  `1`}`2`
}`
\end{lstlisting}
We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
    `1`next{`2`yhat * yhat * 7`1`}
`2`}`
\end{lstlisting}
Once again, we lift the results of specialization into a substitution:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
[zhat|->yhat * yhat * 7]
    `1`next{`2`zhat`1`}`2`
}`
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

@@

\subsubsection{Evaluation and Specialization}
\label{ssec:dynamics}

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(specialization), and the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (residual table).
These judgments are related as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.
In order to avoid duplicating subterms, evaluation collects explicit
substitutions into a single {\em residual table} that is also the output of
evaluation.
\item Specialization sends a stage~\bbtwo\ term to a residual.
\item Reification sends a single residual and a residual table to another residual.  
\item Evaluation depends on specialization for $\next$ terms, which have
stage~\bbtwo\ subterms.
\item Specialization depends on evaluation for $\prev$ and $\pause$, which have
stage~\bbone\ subterms. The residual table produced by evaluation is also
reified here.
\end{itemize}

@@

The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
\bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a {\em residual table} $\xi$ and a partial value $v$. The residual
table implements the explicit substitutions in our example---it maps fresh
stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
residuals. For non-staging features of \lang, $\redonesym$ is essentially
standard call-by-value evaluation, and gathers the subterms' residual tables
into a single one.

@@

At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), specialization simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a residual table $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.
Once specialization reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a residual
table which maps that variable to the residual ($\hat y \mapsto q$).

Within specialization lies a subtle---if perhaps unintuitive---feature.  
Observe that specialization will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
1`next{
  `2`if true 
  then hold{`1`3+4`2`} 
  else prev{`1`spin() (* loops forever *)`2`}`1`
}`
\end{lstlisting}
will fail to evaluate at stage \bbone.
Although it may seem undesirable, this behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

### 2014-11-05 paper/splitting.tex
lots of easy changes in 3 and 4

\ur{Why is there a free variable ``Gamma'' in the figure for splitting
  stage 2?}

@@

\ur{The following paragraph makes no sense to me, especially because
  it does not match Figure 11.  AS far as I can see Figure 11 does not
talk about masking but value splitting.}
We formalize this notion via two \emph{masking} functions (\ref{fig:valSplit}),
$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
a partial value at each stage, essentially by replacing all stage \bbone\
(resp., \bbtwo) components of a partial value by \texttt{()}. The first result
of splitting must then compute the \bbone-mask of the partial value, and the
second result the \bbtwo-mask.

\ur{The following paragraph makes no sense to me, especially because
  it does not match Figure 11.}

@@

Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
future context $\xi$, because the stage \bbtwo\ parts of $v$ contain variables
in $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
structure, and then reify the resulting term in the context $\xi$. This
rewriting process replaces base type constants in $v$ with \texttt{()}, and
$\next~\hat{y}$ with $\hat{y}$. Injection tags are eliminated, because the sum
type is at stage \bbone, so the tag is not relevant at stage \bbtwo. For
functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.

@@

In the splitting rules for stage \bbone\ {\tt case} expressions, we
have enough information at stage \bbone\ to know what branch to take, so there's
no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
taken branch only, and inject the precomputation into a sum type. In the
resumer, we case on that injection, and resume the stage \bbtwo\ portion of the
correct branch.

\crem{split value splitting into two judgments, and call it masking}

\crem{masking figure is missing injections}

\crem{rename future context to residual table}

>>>>

@@

\ur{The following paragraph makes no sense to me, especially because
  it does not match Figure 11.  AS far as I can see Figure 11 does not
talk about masking but value splitting.}
We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
a partial value at each stage, essentially by replacing all stage \bbone\
(resp., \bbtwo) components of a partial value by \texttt{()}. The first result
of splitting must then compute the \bbone-mask of the partial value, and the
second result the \bbtwo-mask.

@@

Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
residual table $\xi$, because the stage \bbtwo\ parts of $v$ contain references
to $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
structure, and then reify the resulting term in the context $\xi$. This
rewriting process replaces base type constants in $v$ with \texttt{()}, and
$\next~\hat{y}$ with $\hat{y}$. Injection tags are eliminated, because the sum
type is at stage \bbone, so the tag is not relevant at stage \bbtwo. For
functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.

@@

In the splitting rules for stage \bbone\ {\tt case} expressions, we
have enough information at stage \bbone\ to know what branch to take, so there's
no need to evaluate both sides. Instead, we evaluate the stage \bbone\ portion of the
taken branch only, and inject the precomputation into a sum type. In the
resumer, we case on that injection, and resume the stage \bbtwo\ portion of the
correct branch.
